 

### 2021-06-28 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd

### In-situ observations
**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in-situ $T_{leaf}$ observations (Figs. `r fig_NEON_vertical`-`r fig_leaf_T`).** 
In closed-canopy forests, *including tropical and temperate broadleaf forests*, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is commonly greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. 
Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007]. 
Under other conditions $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).

>>>>

### In-situ observations
**Following the principles outlined above, vertical microclimate profiles, leaf traits (see "Leaf Traits" section), and $g_s$ (see "Leaf metabolism and thermal responses" section) together explain in-situ $T_{leaf}$ observations (Figs. `r fig_NEON_vertical`-`r fig_leaf_T`).** 
Often, $T_{leaf}$ tracks $T_{air}$, *in large part due to vertical gradients in leaf traits and $g_s$* / ~due to high evapotranspiration~, and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
In closed-canopy forests, *including tropical and temperate broadleaf forests*, leaves experience greater $T_{leaf} - T_{air}$ and higher maximum $T_{leaf}$ in the upper canopies than buffered lower-canopies, sometimes exceeding optima for photosynthesis ($T_{opt}$) and $T_{air}$, because of $g_s$-limitation and high irradiance [@niinemetsShapeLeafPhotosynthetic1999; @doughtyAreTropicalForests2008; @fauset_differences_2018; @mauTemperateTropicalForest2018a; @millerOnlySunlitLeaves2021; @rey-sanchez_spatial_2016; @mauTemperateTropicalForest2018a; @rey-sanchez_spatial_2016; @pauTropicalForestTemperature2018]. 
In contrast, in open forests with lower buffering capacity, lower-canopy $T_{leaf}$ is commonly greater than upper canopy $T_{leaf}$, due to greater light permeability, *vertical air mixing*, and still air in the lower canopy and heating from the ground [Figs. `r fig_NEON_vertical`h, `r SIfig_NEON_all`b,e, @mullerEvidenceEfficientNonevaporative2021; @martin_boundary_1999; @zweifelMiddayStomatalClosure2002; @hadleyInfluenceKrummholzMat1987;@rey-sanchez_spatial_2016]. 
Similarly, trees growing close to the ground can experience greater heat stress in their lower than upper canopies [@hadleyInfluenceKrummholzMat1987; @curtis_intracanopy_2019]. 
Greater $RH$ in closed-inner canopy can also increase $T_{leaf}$ relatively more than upper-canopy $T_{leaf}$ [@dietzVerticalPatternsDuration2007].
